---
title: "Coding Assignment: Distributions, Density, and Overplotting"
subtitle: "A quick recap + practical extensions (bins, bandwidth, alpha, jitter)"
author: "Jared Edgerton"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, plsc498.css]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: ""
---

class: center, middle, inverse, title-slide

.title[
# Coding Assignment: Distributions, Density, and Overplotting
]
.subtitle[
## A quick recap + practical extensions (bins, bandwidth, alpha, jitter)
]
.author[
### Jared Edgerton
]

---

# What you should remember

This is a short recap and extension.

You should already be comfortable with:

- Mapping vs styling
- Choosing a plot for a question (distribution vs relationship)
- Writing code that runs top-to-bottom

Today we extend that with:

- histogram bin choices (`bins`)
- density smoothing choices (`adjust`)
- multiple ways to handle overlapping points (`alpha`, `size`, jitter, bin2d)

---

# Setup

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

set.seed(123)
```

---

# Simulated data

We simulate:

- a skewed variable for histograms / densities
- groups for comparing distributions
- a relationship for scatterplots
- rounding to **force overlap** (so you can see why fixes matter)

```{r}
df <- data.frame(
  id = 1:3500,
  group = sample(c("A", "B"), 3500, replace = TRUE),
  x = rnorm(3500))
df <- df %>%
  mutate(
    income_like = exp(0.7 * x + rnorm(3500, sd = 0.7)),
    y = 1.5 + 0.9 * x + rnorm(3500, sd = 1.2),
    x_round = round(x, 1),
    y_round = round(y, 1)
  )
```

---

# Quick check

```{r}
summary(df$income_like)
summary(df$x)
summary(df$y)
table(df$group)
```

---

# Histogram: default bins (ggplot decides)

Use a histogram to understand **shape**.

```{r, fig.width=6, fig.height=3.5, warning=FALSE, message=FALSE}
ggplot(df, aes(x = income_like)) +
  geom_histogram() +
  theme_classic()
```

Interpretation:
- ggplot chooses binning automatically
- good for a first look
- not something to treat as final by default

---

# Histogram: your bin choice (you decide)

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = income_like)) +
  geom_histogram(bins = 25) +
  theme_classic()
```

Interpretation:
- data unchanged
- only the aggregation changed
- apparent “shape” can change

---

# Histogram: too few

```{r, fig.width=7, fig.height=3.8}
ggplot(df, aes(x = income_like)) +
  geom_histogram(bins = 10) +
  theme_classic()
```

---

# Histogram: too many

```{r, fig.width=7, fig.height=3.8}
ggplot(df, aes(x = income_like)) +
  geom_histogram(bins = 90) +
  theme_classic()
```

Interpretation:
- too few bins can hide structure
- too many bins can create noise
- justify your choice when it matters

---

# Density: a smoothed alternative

A density plot estimates a smooth distribution.

```{r, fig.width=6, fig.height=3.5}
ggplot(df, aes(x = income_like)) +
  geom_density() +
  theme_classic()
```

Interpretation:
- y-axis is **density** (area under the curve = 1)
- not the same as counts

---

# Density: smoothing is a choice (adjust)

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = income_like)) +
  geom_density(adjust = 0.6) +
  theme_classic()
```

---

# Density: smoothing is a choice (adjust)

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = income_like)) +
  geom_density(adjust = 1.8) +
  theme_classic()
```

Interpretation:
- smaller `adjust` = wigglier estimate
- larger `adjust` = smoother estimate

---

# Compare distributions by group: overlay

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = income_like, fill = group)) +
  geom_density(alpha = 0.35) +
  theme_classic()
```

Interpretation:
- overlay works best with few groups
- `alpha` here is helpful because filled shapes overlap

---

# Compare distributions by group: facet

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = income_like)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ group) +
  theme_classic()
```

Interpretation:
- faceting reduces overlap
- makes within-group shape easier to see

---

# Scatterplot: baseline (relationship)

Use a scatterplot when you care about:
- association
- outliers
- nonlinearity

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  theme_classic()
```

---

# Overplotting: force overlap so you can see

Here we plot rounded values.
Many points share the same coordinates.

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = x_round, y = y_round)) +
  geom_point() +
  theme_classic()
```

---

# Overplotting fix 1: alpha + size (styling)

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = x_round, y = y_round)) +
  geom_point(alpha = 0.25, size = 0.8) +
  theme_classic()
```

Interpretation:
- dense regions become darker
- sparse regions remain visible
- alpha/size are **styling** choices when set outside `aes()`

---

# Overplotting fix 2: jitter

Jitter separates coincident points by adding small random noise.

```{r, fig.width=6, fig.height=3.2}
ggplot(df, aes(x = x_round, y = y_round)) +
  geom_point(alpha = 0.25, size = 0.8, position = position_jitter(width = 0.12, height = 0.12)) +
  theme_classic()
```

Interpretation:
- jitter reveals counts at common values
- too much jitter can mislead (it *moves* points)

---

# Overplotting fix 3: 2D binning

Instead of plotting each point, summarize how many fall in each grid cell.

```{r, fig.width=6, fig.height=3.2}
ggplot(df, aes(x = x, y = y)) +
  geom_bin2d() +
  theme_classic()
```

Interpretation:
- you lose individual outliers
- you gain a clear view of dense regions

---

# Another option: density contours

Contours draw “lines of equal point density.”

```{r, fig.width=6, fig.height=3.2}
ggplot(df, aes(x = x, y = y)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_density_2d() +
  theme_classic()
```

Interpretation:
- useful for seeing the *structure* of the cloud
- can get messy with many groups

---

# Mapping vs styling (alpha/size)

**Styling (fixed for all points)**

```r
geom_point(alpha = 0.25, size = 0.8)
```

**Mapping (alpha/size vary by data)**

```r
aes(alpha = income_like)
aes(size = income_like)
```

Early on, prefer styling for readability.

---

# Layers: computed statistics + position

Many geoms compute something automatically.

Example: `geom_bar()` computes counts.

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = group)) +
  geom_bar() +
  theme_classic()
```

---

# Same bar chart, but show proportions

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = group, y = after_stat(count / sum(count)))) +
  geom_bar() +
  theme_classic() +
  labs(y = "Proportion")
```

Interpretation:
- the geometry is the same
- the *statistic you plot* changes the question (counts vs proportions)

---

# Save a figure

Saving your figure is part of making your workflow reproducible.

```{r, fig.width=7, fig.height=4}
p <- ggplot(df, aes(x = x, y = y)) +
  geom_point(alpha = 0.2, size = 0.7) +
  theme_classic()

p

ggsave("scatter_alpha.png", plot = p, width = 7, height = 4, units = "in")
```
