---
title: "Coding Assignment: Histograms, Scatterplots, and Overplotting"
subtitle: "A quick recap + two practical extensions (bins, alpha, size)"
author: "Jared Edgerton"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, plsc498.css]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: ""
---


# What you should remember

This is a short recap and extension.

You should already be comfortable with:

- Mapping vs styling
- Choosing a plot for a question (distribution vs relationship)
- Writing code that runs top-to-bottom

Today we extend that with:

- histogram bin choices
- using `alpha` and `size` to handle overplotting

---

# Setup

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

set.seed(123)
```

---

# Simulated data (designed for these plots)

We simulate:
- a skewed variable for histograms
- a noisy relationship for scatterplots
- enough points to create overplotting

```{r}
df <- data.frame(
  id = 1:2000,
  group = sample(c("A", "B"), 2000, replace = TRUE),
  x = rnorm(2000)
)
df <- df %>%
  mutate(
    # positive + skewed (common in “real-world” quantities)
    income_like = exp(0.7 * x + rnorm(2000, sd = 0.6)),
    # outcome related to x, with noise
    y = 2 + 0.8 * x + rnorm(2000, sd = 1.2)
  )
```

---

# Quick check

```{r}
summary(df$income_like)
summary(df$x)
summary(df$y)
table(df$group)
```

---

# Histogram: default bins (ggplot decides)

Use a histogram to understand **shape**.

```{r fig.width=5, fig.height=3, warning=FALSE, message=FALSE}
ggplot(df, aes(x = income_like)) +
  geom_histogram() +
  theme_classic()
```

Interpretation:
- ggplot chooses a bin width automatically
- good for a first look
- not something to treat as final by default

---

# Histogram: your bin choice (you decide)

```{r fig.width=7, fig.height=4}
ggplot(df, aes(x = income_like)) +
  geom_histogram(bins = 30) +
  theme_classic()
```

Interpretation:
- data unchanged
- only the aggregation changed
- apparent “shape” can change

---

# Histogram: too few 

```{r fig.width=7, fig.height=4}
ggplot(df, aes(x = income_like)) +
  geom_histogram(bins = 10) +
  theme_classic()
```

---

# Histogram: too many bins


```{r fig.width=7, fig.height=4}
ggplot(df, aes(x = income_like)) +
  geom_histogram(bins = 80) +
  theme_classic()
```

Interpretation:
- too few bins can hide structure
- too many bins can create noise
- justify your choice when it matters

---

# Scatterplot: baseline (relationship)

Use a scatterplot when you care about:
- association
- outliers
- nonlinearity

```{r fig.width=5, fig.height=3}
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  theme_classic()
```

---

# Overplotting: why it happens

When many points overlap:
- dense regions become solid blobs
- rare regions can disappear
- you lose information

Two simple tools:
- `alpha` (transparency)
- `size` (point size)

---

# Scatterplot: alpha (transparency) intervention

```{r fig.width=5, fig.height=3}
ggplot(df, aes(x = x, y = y)) +
  geom_point(alpha = 0.2) +
  theme_classic()
```

Interpretation:
- dense regions become darker
- sparse regions remain visible
- alpha is a styling choice (not a mapping)

---

# Scatterplot: size intervention

```{r fig.width=5, fig.height=3}
ggplot(df, aes(x = x, y = y)) +
  geom_point(size = 0.7, alpha = 0.3) +
  theme_classic()
```

Interpretation:
- smaller points reduce clutter
- combined with alpha, this is a strong default for many scatterplots

---

# Mapping vs styling (alpha/size)

**Styling (fixed for all points)**

```r
geom_point(alpha = 0.2, size = 0.7)
```

**Mapping (alpha/size vary by data)**

```r
aes(alpha = income_like)
aes(size = income_like)
```

Early on, prefer styling for readability.

---

# Optional: mapping size to a variable (use with care)

Mapping size can be useful, but it adds complexity. Start with a small range of sizes.

```{r fig.width=4, fig.height=2.5}
ggplot(df, aes(x = x, y = y, size = income_like)) +
  geom_point(alpha = 0.25) +
  theme_classic()
```

Interpretation:
- size encodes a third variable
- can be hard to read precisely
- can create “proportional ink” issues if overused


